Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-10-19 paper.tex
Len first contrib

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools,enumitem}
\usepackage{algorithm}
\usepackage{algorithmic}
%\documentstyle[nips14submit_09,times,art10]{article} % For LaTeX 2.09

@@

\title{Over-relaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport}

%
\author{
Alexis THIBAULT\\
\'Ecole Normale Sup\'erieure\\
Paris, France\\
\texttt{alexis.thibault@ens.fr}
 \And
L\'enaic CHIZAT\\
\'Ecole Normale Sup\'erieure\\ Paris Dauphine (PSL research University)\\
Paris, France\\
\texttt{lenaic.chizat@ens.fr}
 \AND
Charles DOSSAL\\
INSA Toulouse\\
Institut de Math\'ematiques de Toulouse\\
Toulouse,  France\\
\texttt{dossal@insa-toulouse.fr}
\And 
Nicolas PAPADAKIS\\
CNRS, Institut de Math\'ematiques de Bordeaux\\
Talence, France\\
\texttt{nicolas.papadakis@math.u-bordeaux.fr}
} 
% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

@@

\begin{abstract}
This article describes a method for quickly calculating the solution to the regularized optimal transport problem. It generalizes and improves upon the widely-used iterative Bregman projections algorithm (or Sinkhorn--Knopp algorithm). 
The idea is to over-relax the Bregman projection operators, allowing for faster convergence. In practice this corresponds to elevating the diagonal scaling factors to a given power, at each step of the algorithm.
In this paper, we propose a simple method for establishing the global convergence of the over-relaxed Sinkhorn--Knopp algorithm, by ensuring the decrease of a Lyapunov function at each step.
An inexpensive adaptive choice of over-relaxation parameter based on the Lyapunov function is constructed.
We also suggest a heuristic to choose a suitable asymptotic over-relaxation parameter.
\end{abstract}

\section{Introduction}
Optimal Transport is an efficient and flexible tool to compare two probability distributions which has been popularized in the computer vision community in the context of discrete histograms \cite{Rubner2000}. The introduction of entropic regularization in \cite{cuturi13} has made possible the use of the fast Sinkhorn--Knopp algorithm \cite{sinkhorn64}   scaling with high dimensional data. 
Regularized optimal transport have thus been intensively used  in  Machine Learning with applications such as   Geodesic PCA \cite{seguy2015principal}, domain adaptation \cite{2015arXiv150700504C}, data fitting \cite{2015arXiv150605439F},  training of Boltzmann Machine \cite{NIPS2016_6248}  or dictionary learning \cite{Rolet2016,2017arXiv170801955S}.

The computation of optimal transport between two data relies on the estimation of an optimal transport matrix, the entries of which represent the quantity of mass transported between  data locations. 
Regularization of optimal transport with strictly convex regularizers \cite{cuturi13, dessein2016}  nevertheless involves a spreading of the mass. Hence, for particular purposes such as color interpolation \cite{Rabin2014} or gradient flow \cite{2016arXiv160705816C}, it is  necessary  to consider very small regularization of the problem.
In this setting,  the regularized transport problem can be ill-conditioned. This is the issue  we want to tackle here.
Before going into further details, we now briefly introduce the main notations and concepts used all along this article.

\subsection{Discrete Optimal Transport}
We consider two discrete probability measures $\mu^k \in \IR_{+*}^{n_k}$.
Let us define the two following linear operators
\begin{align*}
A_1 &: \begin{cases}
\IR^{n_1 n_2} \rightarrow \IR^{n_1} \\
(A_1 x)_i = \sum_j x_{i,j}
\end{cases} &
A_2 &: \begin{cases}
\IR^{n_1 n_2} \rightarrow \IR^{n_2}\\
(A_2 x)_j = \sum_i x_{i,j},
\end{cases}
\end{align*}
as well as the linear  constraint sets
\begin{align*}
\Ccal_k &= \left\{ \gamma\in\IR^{n_1 n_2} \mid A_k \gamma = \mu^k \right\}.
\end{align*}
Given a cost matrix $c$, where $c_{ij}$ represents the cost of moving mass $\mu^1_i$ to $\mu^2_j$,  the optimal transport problem corresponds to the estimation of an optimal transport matrix $\gamma$ solution of:
$$\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_+} \langle c,\gamma\rangle:=\sum_{i,j}c_{i,j}\gamma_{i,j}.$$
This is a linear programing problem whose resolution is unmanageable for large scale problems.

@@

In \cite{cuturi13}, it has been proposed to regularize this problem by adding a strictly convex entropy regularization:
\begin{equation}\label{ROT}
\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_{+*}}K^\epsilon(\gamma) := \scal{c}{\gamma} 
+ \epsilon KL(\gamma,\mathbf{1})
,\end{equation}
with $\epsilon>0$, $\mathbf{1}$ is the matrix of size $n_1\times n_2$ full of ones and the Kullback-Leibler divergence is:
\begin{equation}\label{KL}
KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}.
\end{equation}
It was shown in \cite{benamou15}  that the regularized optimal transport matrix $\gamma^*$, which is the unique minimizer of problem \eqref{ROT},  is the Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ onto $\Ccal_1 \cap \Ccal_2$:
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma)= P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon}),
\end{equation}
where $P_{\Ccal}$ is the  Bregman projection onto $\Ccal$ defined as:
\[
P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} KL(\gamma,\xi).
\]

\subsection{The Sinkhorn--Knopp algorithm}
Iterative Bregman projections onto $\Ccal_1$ and $\Ccal_2$ converge to a point in the intersection $\Ccal_1 \cap \Ccal_2$ \cite{bregman67}. Hence, the so-called Sinkhorn--Knopp algorithm \cite{sinkhorn64} that performs alternate Bregman projections, can be considered to compute the regularized  transport matrix:
\begin{align*}
\gamma^0 &= e^{-c/\epsilon} &
\gamma^{l+1} = P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l)),
\end{align*}
and we have 
$\lim_{l\rightarrow +\infty} \gamma^l = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0) = \gamma^*.$

In the discrete setting,  projections actually correspond to diagonal scalings.
\begin{align}\label{scaling}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &
a &=  {\mu^1}\oslash{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &
b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
\end{align}
where $\oslash$ is the pointwise division. 
To compute numerically the solution  one simply has to store $(a^l, b^l)\in\IR^{n_1+n_2}$ and to iterate
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} 
\end{align*}
where the transport matrix can be recovered as :
$\gamma^l = \diag(a^l) \gamma^0 \diag(b^l).$ 
Efficient parallel computations can be considered \cite{cuturi13} and one can even reach real-time computation for large scale problem for certain class of cost matrices $c$ \cite{Solomon2015}. 
For small values of the parameter $\epsilon$, numerical issues can arise and a stabilization of the algorithm is necessary \cite{2016arXiv160705816C}.
The convergence of the process can nevertheless be very slow  in the setting $\epsilon\to 0$.
\subsection{Related works }
The introduction of relaxation variables through heavy ball approaches \cite{POLYAK19641} has recently gained in popularity  to speed up the convergence of algorithms optimizing convex \cite{2014arXiv1412.7457G} or non convex \cite{Zavriev1993,2016arXiv160609070O} problems. Such schemes have also been empirically considered to accelerate the SK algorithm  in \cite{peyre2016quantum,2017arXiv170801955S}. The convergence of these algorithms is nevertheless not studied yet in the context of regularized optimal transport.

\subsection{Overview and contributions}
In this paper, we consider an over-relaxation scheme designed to accelerate the Sinkhorn--Knopp algorithm. We first present and show the convergence of our algorithm in section 2. In section 3, we analyze the local convergence rate of the algorithm to justify the acceleration.
We finally demonstrate numerically  in Section 4 the good behavior of our method, where larger accelerations are observed for decreasing values of $\epsilon$.

\section{Over-relaxed Sinkhorn--Knopp algorithm}

As illustrated in Figure \ref{alternate_projections} (a-b), the SK algorithm, that  performs alternate Bregman projections onto the affine sets $\Ccal_1$ and $\Ccal_2$, can be very slow when $\epsilon\to 0$. The idea developped in this paper is to perform over-relaxed projections in order to accelerate the process, as displayed in Figure \ref{alternate_projections} (c).

\begin{figure}[ht!]
\begin{center}
\begin{tabular}{ccc}
\input{schema_a.tex}&
\input{schema_b.tex}&
\input{schema_c.tex}\\
(a)&(b)&(c)
\end{tabular}
\caption{\label{alternate_projections} The trajectory of $\gamma^l$ given by the SK algorithm is illustrated for decreasing values of $\epsilon$ in (a) and (b). In order to accelerate the convergence rate, over-relaxed projections are here considered (c).}
\end{center}
\end{figure}
\subsection{Over-relaxed projections}

Let us define the $\omega$-over-relaxed projection operator by:
\begin{equation}\label{eq:def_or_proj}
\log P^\omega_{\Ccal_k}(\gamma) = (1-\omega) \log \gamma + \omega \log P_{\Ccal_k}(\gamma),
\end{equation}
where the logarithm is taken coordinate-wise.
Note that $P_{\Ccal_k}^0$ is the identity, $P_{\Ccal_k}^1 = P_{\Ccal_k}$ is the standard Bregman projection, and $P_{\Ccal_k}^2$ is an involution (this uses the fact that $\Ccal_k$ is an affine subspace).

{\color{blue} \begin{proposition} 
State here that the fixed points of   $\gamma^{l+1} =P^\theta_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma^l))$ are the same than $\gamma^{l+1} =P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l))$?
\end{proposition}}
\subsection{Lyapunov function}
Let $\gamma^*$ denote the solution of the regularized OT problem.
The function $F$ is defined as:
\begin{equation}\label{eq:lyapunov_function}
F(\gamma) = KL(\gamma^*, \gamma)
\end{equation}
It shall be used as a Lyapunov function, so that $F(\gamma^{k+1}) < F(\gamma^k)$ as long as the process has not converged.

We argue that it makes sense to consider this Lyapunov function.
Obviously, it is closely related to the notion of Bregman projection that is used throughout the algorithm.
But it is also easy to calculate its decrease, as shown in lemma \ref{lemma:lyapunov_decrease}.
Moreover, in practice it allows the use of a wide range of over-relaxation parameters.

\begin{lemma} \label{lemma:KL_compact}
	For any $M \in \IR_+^*$, the sublevel set $K := \left\{ \gamma \mid F(\gamma) \le M \right\}$ is compact.
\end{lemma}
\begin{proof}
	First note that K is closed in $\IR_{+*}^{n_1 n_2}$, by continuity of $F$.
	For $\gamma \in K$, one has:
	\begin{align*}
	F(\gamma) &= \scal{\gamma^*}{\log \gamma^* - 1} + \scal{\gamma}{1} - \scal{\gamma^*}{\log \gamma}\\
	\sum_{i,j} \gamma_{i,j} - \gamma^*_{i,j} \log \gamma_{i,j} &\le M - \scal{\gamma^*}{\log \gamma^* - 1} .
	\end{align*}
	By concavity, the logarithm admits a simple upper bound: $\log \gamma_{i,j} \le \frac{\gamma_{i,j}}{e}$. Moreover, since $\gamma^*$ is a solution of the problem, each of its coordinates is at most 1.
	\begin{align*}
	\sum_{i,j} \gamma_{i,j} &\le \frac{M - \scal{\gamma^*}{\log \gamma^*-1}}{1-\frac{1}{e}} =: M'.
	\end{align*}
	Therefore each coordinate $\gamma_{i,j}$ is bounded above by $M'$.

	Let us show that the coordinates are bounded below by a strictly positive number; from which it can be deduced that $K$ is bounded and closed in $\IR^{n_1 n_2}$, and therefore compact.
	\begin{align*}
	F(\gamma)
	&\ge \scal{\gamma^*}{\log \gamma^*} - \sum_{i,j} \gamma^*_{i,j} \log(\gamma_{i,j}) - 1\\
	&\ge \scal{\gamma^*}{\log \gamma^*} - \gamma^*_{i_0,j_0} \log(\gamma_{i_0,j_0}) - 1 - \log M',
	\end{align*}
	where the last equation derives from $\log(y_{i,j}) \le \log M'$, and $\sum_{(i,j)\neq(i_0,j_0)} \gamma^*_{i,j} \le 1$.
	Finally:
	\[
	\log(\gamma_{i_0,j_0}) \ge \frac{\scal{\gamma^*}{\log \gamma^*} - M - \log M' - 1}{\gamma^*_{i_0,j_0}},
	\]
	and $K$ is thus compact.
\end{proof}

Note that the difference $F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma))$ may be calculated without knowing $\gamma^*$, as shown by the following lemma.
\begin{lemma}\label{lemma:lyapunov_decrease}
	Take $\gamma$ in $\IR^{mn}_{+*}$. The decrease in value of the Lyapunov function can be calculated with the following formula:
	\begin{equation} \label{eq:kl_diff_scal}
	F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma)) = 
	\scal{\mu^k}{\varphi_\omega \left(\frac{A_k \gamma}{\mu^k}\right)},
	\end{equation}
	where
	\begin{equation}
	\varphi_\omega(x) = x(1-x^{-\omega}) - \omega \log x
	\end{equation}
	is a real function, applied coordinate-wise.
\end{lemma}
\begin{proof}
From \eqref{KL}, we have $F(\gamma^1)-F(\gamma^2)= \sum_{i,j}\left(\gamma^*_{i,j}\log(\gamma^2_{i,j}/\gamma^1_{i,j})+\gamma^1_{i,j}-\gamma^2_{i,j}\right)$. The result can then be deduced from relations \eqref{eq:def_or_proj} and \eqref{scaling}.
\end{proof}
Thus,  the decrease in value of the function $F$ for a over-relaxed projection associated to $\Ccal_k$ is cheap to estimate, since its computational cost is  linear with respect to the dimension of data $\mu^k$. In Figure \ref{phi_omega}, we illustrate the functions  $\varphi_\omega$ for several values of $\omega$.
Notice that for the SK algorithm, which corresponds to $\omega=1$, the function $\varphi_\omega$ is always nonnegative. For other values $1\le\omega<2$, it is nonnegative for arguments close to 1.
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=11cm]{phi_omega.png}
\caption{\label{phi_omega} Plot of functions $\varphi_\omega(x)$ for different values $1\leq \omega\leq 2$ and   $x\in[10^{-2}, 10^1]$.}
\end{center}
\end{figure}
\subsection{Proposed algorithm}

\begin{theorem}\label{thm:algo}
	Take two continuous functions $\theta_1,\theta_2$ such that
	\begin{equation}\label{eq:cond_theta_k}
	\forall \gamma \in \IR_{+*}^{n_1 n_2},\quad
	F(P_{\Ccal_k}^{\theta_k(\gamma)}(\gamma)) \le F(\gamma) ,
	\end{equation}
	where the inequality is strict whenever $\gamma \notin \Ccal_k$.
	Consider the sequence defined by:
	\begin{align*}
	\gamma^0 &= e^{-c/\epsilon} \\
	\tilde{\gamma}^{l+1} &= P_{\Ccal_1}^{\theta_1(\gamma^l)}(\gamma^l) \\
	\gamma^{l+1} &= P_{\Ccal_2}^{\theta_2(\tilde{\gamma}^{l+1})}(\tilde{\gamma}^{l+1}).
	\end{align*}
	Then the sequence $(\gamma^l)$ converges to $\gamma^*$.
\end{theorem}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools,enumitem}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
%\documentstyle[nips14submit_09,times,art10]{article} % For LaTeX 2.09

@@

\title{Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport}

%
\author{
Alexis THIBAULT\\
\'Ecole Normale Sup\'erieure\\
Paris, France\\
\texttt{alexis.thibault@ens.fr}
 \And
L\'ena\"ic CHIZAT\\
\'Ecole Normale Sup\'erieure\\ Paris Dauphine (PSL research University)\\
Paris, France\\
\texttt{lenaic.chizat@ens.fr}
 \AND
Charles DOSSAL\\
INSA Toulouse\\
Institut de Math\'ematiques de Toulouse\\
Toulouse,  France\\
\texttt{dossal@insa-toulouse.fr}
\And 
Nicolas PAPADAKIS\\
CNRS, Institut de Math\'ematiques de Bordeaux\\
Talence, France\\
\texttt{nicolas.papadakis@math.u-bordeaux.fr}
} 
% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

@@

\begin{abstract}
This article describes a method for quickly computing the solution to the regularized optimal transport problem. It generalizes and improves upon the widely-used iterative Bregman projections algorithm (or Sinkhorn--Knopp algorithm). 
The idea is to overrelax the Bregman projection operators, allowing for faster convergence. In practice this corresponds to elevating the diagonal scaling factors to a given power, at each step of the algorithm.
We propose a simple method for establishing global convergence by ensuring the decrease of a Lyapunov function at each step.
An adaptive choice of overrelaxation parameter based on the Lyapunov function is constructed.
We also suggest a heuristic to choose a suitable asymptotic over-relaxation parameter, based on a local convergence analysis. Our numerical experiments show a gain in convergence speed by an order of magnitude in certain regimes.
\end{abstract}

\section{Introduction}
Optimal Transport (OT) is an efficient and flexible tool to compare two probability distributions which has been popularized in the computer vision community in the context of discrete histograms \cite{Rubner2000}. The introduction of entropic regularization in \cite{cuturi13} has made possible the use of the fast Sinkhorn--Knopp algorithm \cite{sinkhorn64}   scaling with high dimensional data. 
Regularized optimal transport have thus been intensively used  in  Machine Learning with applications such as   Geodesic PCA \cite{seguy2015principal}, domain adaptation \cite{2015arXiv150700504C}, data fitting \cite{2015arXiv150605439F},  training of Boltzmann Machine \cite{NIPS2016_6248}  or dictionary learning \cite{Rolet2016,2017arXiv170801955S}.

The computation of optimal transport between two data relies on the estimation of an optimal transport matrix, the entries of which represent the quantity of mass transported between  data locations. 
Regularization of optimal transport with strictly convex regularizers \cite{cuturi13, dessein2016}  nevertheless involves a spreading of the mass. Hence, for particular purposes such as color interpolation \cite{Rabin2014} or gradient flow \cite{2016arXiv160705816C}, it is  necessary  to consider very small regularization of the problem.
In this setting,  the regularized transport problem can be ill-conditioned and Sinkhorn's algorithm converges slowly. This is the issue  we want to tackle here.
Before going into further details, we now briefly introduce the main notations and concepts used all along this article.

\subsection{Discrete optimal transport}
We consider two discrete probability measures $\mu^k \in \IR_{+*}^{n_k}$.
Let us define the two following linear operators
\begin{align*}
A_1 &: \begin{cases}
\IR^{n_1 n_2} \rightarrow \IR^{n_1} \\
(A_1 x)_i = \sum_j x_{i,j}
\end{cases} &
A_2 &: \begin{cases}
\IR^{n_1 n_2} \rightarrow \IR^{n_2}\\
(A_2 x)_j = \sum_i x_{i,j},
\end{cases}
\end{align*}
as well as the affine constraint sets
\begin{align*}
\Ccal_k &= \left\{ \gamma\in\IR^{n_1 n_2} \mid A_k \gamma = \mu^k \right\}.
\end{align*}
Given a cost matrix $c$, where $c_{ij}$ represents the cost of moving mass $\mu^1_i$ to $\mu^2_j$,  the optimal transport problem corresponds to the estimation of an optimal transport matrix $\gamma$ solution of:
$$\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_+} \langle c,\gamma\rangle:=\sum_{i,j}c_{i,j}\gamma_{i,j}.$$
This is a linear programming problem whose resolution becomes intractable for large problems.

@@

In \cite{cuturi13}, it has been proposed to regularize this problem by adding a strictly convex entropy regularization:
\begin{equation}\label{ROT}
\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_{+}}K^\epsilon(\gamma) \coloneqq \scal{c}{\gamma} 
+ \epsilon \KL(\gamma,\mathbf{1})
,\end{equation}
with $\epsilon>0$, $\mathbf{1}$ is the matrix of size $n_1\times n_2$ full of ones and the Kullback-Leibler divergence is
\begin{equation}\label{KL}
\KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}
\end{equation}
with the convention $0 \log 0= 0$. It was shown in \cite{benamou15}  that the regularized optimal transport matrix $\gamma^*$, which is the unique minimizer of problem \eqref{ROT},  is the Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ (here and in the sequel, exponentiation is meant entrywise) onto $\Ccal_1 \cap \Ccal_2$:
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma)= P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon}),
\end{equation}
where $P_{\Ccal}$ is the  Bregman projection onto $\Ccal$ defined as
\[
P_{\Ccal}(\xi) \coloneqq \argmin_{\gamma \in \Ccal} \KL(\gamma,\xi).
\]

\subsection{Sinkhorn--Knopp algorithm}
Iterative Bregman projections onto $\Ccal_1$ and $\Ccal_2$ converge to a point in the intersection $\Ccal_1 \cap \Ccal_2$ \cite{bregman67}. Hence, the so-called Sinkhorn--Knopp algorithm (SK) \cite{sinkhorn64} that performs alternate Bregman projections, can be considered to compute the regularized  transport matrix:
\begin{align*}
\gamma^0 &= e^{-c/\epsilon} &
\gamma^{l+1} = P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l)),
\end{align*}
and we have 
$\lim_{l\rightarrow +\infty} \gamma^l = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0) = \gamma^*.$
%
In the discrete setting, these projections correspond to diagonal scalings of the input:
\begin{align}\label{scaling}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &\text{with}\quad
a &=  {\mu^1}\oslash{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &\text{with}\quad
b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
\end{align}
where $\oslash$ is the pointwise division. 
To compute numerically the solution one simply has to store $(a^l, b^l)\in\IR^{n_1}\times \IR^{n_2}$ and to iterate
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} .
\end{align*}
We then have $\gamma^l = \diag(a^l) \gamma^0 \diag(b^l).$

Another way to interpret the SK algorithm is as an alternate maximization algorithm on the dual of the regularized optimal transport problem. The dual problem of \eqref{ROT} is
\begin{equation}\label{DROT}
\max_{\substack{\alpha\in \IR^n\\\beta\in \IR^m}}\; E(\alpha,\beta) \coloneqq \langle \alpha,\mu^1\rangle+\langle \beta,\mu^2\rangle-\epsilon\sum_{i,j}e^{(\alpha_i+\beta_j-c_{i,j})/\epsilon}.
\end{equation}
The function $E$ is concave, continuously differentiable and admits a maximizer, so alternate maximization converges and we recover, for $a_i=e^{\alpha_i/\epsilon}$, $b_j=e^{\beta_j/\epsilon}$ and $\gamma^0_{i,j}=e^{-c_{i,j}/\epsilon}$:
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} 
\end{align*}

Efficient parallel computations can be considered \cite{cuturi13} and one can even reach real-time computation for large scale problem for certain class of cost matrices $c$ \cite{Solomon2015}. 
For small values of the parameter $\epsilon$, numerical issues can arise and a stabilization of the algorithm is necessary \cite{2016arXiv160705816C}.
The convergence of the process can nevertheless be very slow when $\epsilon$ is small.

\subsection{Overview and contributions}
In this paper, we consider an overrelaxation scheme designed to accelerate the Sinkhorn--Knopp algorithm. We first present and show the convergence of our algorithm in section 2. In section 3, we analyze the local convergence rate of the algorithm to justify the acceleration.
We finally demonstrate numerically  in Section 4 the good behavior of our method, where larger accelerations are observed for decreasing values of $\epsilon$.

\subsection{Related works }
The introduction of relaxation variables through heavy ball approaches \cite{POLYAK19641} has recently gained in popularity  to speed up the convergence of algorithms optimizing convex \cite{2014arXiv1412.7457G} or non convex \cite{Zavriev1993,2016arXiv160609070O} problems. Such schemes have also been empirically considered to accelerate the SK algorithm  in \cite{peyre2016quantum,2017arXiv170801955S}. The convergence of these algorithms is nevertheless not studied yet in the context of regularized optimal transport.

\section{Overrelaxed Sinkhorn--Knopp algorithm}

As illustrated in Figure \ref{alternate_projections} (a-b), the SK algorithm, that  performs alternate Bregman projections onto the affine sets $\Ccal_1$ and $\Ccal_2$, can be very slow when $\epsilon\to 0$. The idea developped in this paper is to perform overrelaxed projections in order to accelerate the process, as displayed in Figure \ref{alternate_projections} (c).

\begin{figure}[ht!]
\centering
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_a.tex}
                    \subcaption{}\label{fig:schema_a}
\end{minipage}%
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_b.tex}
                    \subcaption{}\label{fig:schema_b}
\end{minipage}%
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_c.tex}
                    \subcaption{}\label{fig:schema_c}
\end{minipage}%
\caption{\label{alternate_projections} The trajectory of $\gamma^l$ given by the SK algorithm is illustrated for decreasing values of $\epsilon$ in (a) and (b). Overrelaxed projection typically accelerate the convergence rate.}
\end{figure}

\iffalse
\begin{figure}[ht!]
\begin{center}
\begin{tabular}{ccc}
\input{schema_a.tex}&
\input{schema_b.tex}&
\input{schema_c.tex}\\
(a)&(b)&(c)
\end{tabular}
\caption{\label{alternate_projections} The trajectory of $\gamma^l$ given by the SK algorithm is illustrated for decreasing values of $\epsilon$ in (a) and (b). In order to accelerate the convergence rate, over-relaxed projections are here considered (c).}
\end{center}
\end{figure}
\fi

\subsection{Overrelaxed projections}

We define the $\omega$-overrelaxed projection operator $P^\omega_{\Ccal_k}$ as
\begin{equation}\label{eq:def_or_proj}
\log P^\omega_{\Ccal_k}(\gamma) = (1-\omega) \log \gamma + \omega \log P_{\Ccal_k}(\gamma),
\end{equation}
where the logarithm is taken coordinate-wise. Note that $P_{\Ccal_k}^0$ is the identity, $P_{\Ccal_k}^1 = P_{\Ccal_k}$ is the standard Bregman projection and $P_{\Ccal_k}^2$ is an involution (in particular because $\Ccal_k$ is an affine subspace). A naive algorithm would then consist in iteratively applying $P^\omega_{\Ccal_2}\circ P^\omega_{\Ccal_1}$ for some choice of $\omega$. This algorithm, while it may behave well in practice, has no convergence guarantee---it sometimes does not convergence even for reasonable values of $\omega$. Our goal in this section is to modify this algorithm to make it robust and garantee convergence.

Duality gives another point of view on the iterative overrelaxed Bregman projections. They indeed correspond to a successive overelaxation (SOR) algorithm on the dual objective $E$. This can be seen by using the relationships gievn after equation~\eqref{DROT}:
\begin{align}\label{SORdual}
\alpha^{\ell+1} &= (1-\omega)\alpha^{\ell} + \omega \arg \max_\alpha E(\alpha,\beta^{\ell})\\
\beta^{\ell + 1}&=(1-\omega) \beta^{\ell} + \omega \arg \max_\beta E(\alpha^{\ell+1},\beta).
\end{align}

{\color{blue} \begin{proposition} 
State here that the fixed points of   $\gamma^{l+1} =P^\theta_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma^l))$ are the same than $\gamma^{l+1} =P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l))$? lenaic: what is exactly needed? I can think of it tomorow.
\end{proposition}}
\subsection{Lyapunov function}
Convergence of the successive overrelaxed projection cannot be guaranteed in general. In order to derive a robust algorithm with provable convergence, we introduce the Lyapunov function 
\begin{equation}\label{eq:lyapunov_function}
F(\gamma) = \KL(\gamma^*, \gamma),
\end{equation}
where $\gamma^*$ denotes the solution of the regularized OT problem. We will use this function to enforce the strict descent criterion $F(\gamma^{k+1}) < F(\gamma^k)$ as long as the process has not converged.
%
The choice of \eqref{eq:lyapunov_function} as a Lyapunov function is of course related to the fact that Bregman projections are used throughout the algorithm. Further, we will show (Lemma \ref{lemma:lyapunov_decrease}) that its decrease is very easy to compute and this descent criterion still allows enough freedom in the choice of the overrelaxation parameter.

Crucial properties of this Lyapunov function are gathered in the next lemma.
\begin{lemma} \label{lemma:lyapunov_decrease}
	For any $M \in \IR_+^*$, the sublevel set $\left\{ \gamma \mid F(\gamma) \le M \right\}$ is compact.
%\end{lemma}
%\begin{lemma}
	Moreover, for $\gamma$ in $\IR^{mn}_{+*}$, the decrease of the Lyapunov function after an overrelaxed projection can be computed as
	\begin{equation} \label{eq:kl_diff_scal}
	F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma)) = 
	\scal{\mu^k}{\varphi_\omega \left((A_k \gamma) \oslash \mu^k \right)},
	\end{equation}
	where
	\begin{equation}
	\varphi_\omega(x) = x(1-x^{-\omega}) - \omega \log x
	\end{equation}
	is a real function, applied coordinate-wise.
\end{lemma}

\begin{proof}
The fact that the Kullback-Leibler divergence is jointly lower semicontinuous implies in particular that $K$ is closed. Moreover, $K\subset \mathbb{R}^{n_1\times n_2}_+$ is bounded because $F$ is the sum of nonnegative, coercive functions of each component of its argument $\gamma$.
%\end{proof}
%\begin{proof}
%
Formula \eqref{eq:kl_diff_scal} comes the expression $F(\gamma^1)-F(\gamma^2)= \sum_{i,j}\left(\gamma^*_{i,j}\log(\gamma^2_{i,j}/\gamma^1_{i,j})+\gamma^1_{i,j}-\gamma^2_{i,j}\right)$ and the relations \eqref{eq:def_or_proj} and \eqref{scaling}.
\end{proof}

\iffalse
\begin{proof}
	First note that K is closed in $\IR_{+*}^{n_1 n_2}$, by continuity of $F$.
	For $\gamma \in K$, one has:
	\begin{align*}
	F(\gamma) &= \scal{\gamma^*}{\log \gamma^* - 1} + \scal{\gamma}{1} - \scal{\gamma^*}{\log \gamma}\\
	\sum_{i,j} \gamma_{i,j} - \gamma^*_{i,j} \log \gamma_{i,j} &\le M - \scal{\gamma^*}{\log \gamma^* - 1} .
	\end{align*}
	By concavity, the logarithm admits a simple upper bound: $\log \gamma_{i,j} \le \frac{\gamma_{i,j}}{e}$. Moreover, since $\gamma^*$ is a solution of the problem, each of its coordinates is at most 1.
	\begin{align*}
	\sum_{i,j} \gamma_{i,j} &\le \frac{M - \scal{\gamma^*}{\log \gamma^*-1}}{1-\frac{1}{e}} =: M'.
	\end{align*}
	Therefore each coordinate $\gamma_{i,j}$ is bounded above by $M'$.
